---
title: "R Notebook"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

**Work breakdown structure estimating the efforts of each team
member:**\
Task 1-2 did Anton Deputat and he did 33.33% of work.\
Task 3 + general conclusions did Bohdan Kolodchak and he did 33.33% of
work.\
Task 4 did Mariia Hamaniuk and she did 33.33% of work.

**Problem 1.** $H_0$ : $µ_1$ = $µ_2$ vs. $H_1$ : $µ_1$ ̸= $µ_2$;

$\sigma^2_1$ = $\sigma^2_2$ =1

**R Code:**

```{r}

```

**The statistics obtained (like sample mean or anything else you use to
complete the task):**

**Conclusions:**

**Problem 2.** $H_0$ : $\sigma^2_1$ = $\sigma^2_2$ vs. $H_1$ :
$\sigma^2_1$ \> $\sigma^2_2$ ; $µ_1$ and $µ_2$ are unknown. (Hint: this
is the f-test; read the details in Ross, p. 321–323)

**the statistics obtained (like sample mean or anything else you use to
complete the task):**

**Conclusions:**

**Problem 3.** Using Kolmogorov–Smirnov test in R, check if (a)
${ x_k }_{k=1}^{100}$ are normally distributed (with parameters
calculated from the sample); (b)${|x_k|}_{k=1}^{100}$ are exponentially
distributed with λ = 1; (c) ${ x_k }_{k=1}^{100}$ and
${ y_l }_{l=1}^{50}$ have the same distribution

```{r}

```

Explain the main idea behind the KS test and comment on the outcomes of
the test:

Note: ${ x_k }_{k=1}^{100}$ means a set of a hundred xk with
corresponding indices here, not their fractional part:

**the statistics obtained (like sample mean or anything else you use to
complete the task):**

**Conclusions:**

\

**Problem 4.** In this task you’ll practice fitting the regression line
to some real-life features and analyzing the results. The file data.csv
contains data on students, specifically their study time and
corresponding marks. Your tasks are as follows: (a) Create a scatter
plot of Marks vs. Study Time and provide brief comments;

```{r}

```

(b) Fit a linear regression model using marks as the dependent variable
    and study time as the independent variable.

```{r}

```

Explain shortly the process of deriving the regression equation:

(c) Evaluate the goodness-of-fit for the fitted line;

```{r}

```

**the statistics obtained (like sample mean or anything else you use to
complete the task)**:

**Conclusions:**
